---
title: "AD Genetics Descriptives"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
author:
    - "Shea Andrews"
params:
    rmd: "report.Rmd"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(purrr)
library(janitor)
library(tibble)
library(glue)
library(stringr)
library(readr)
library(forcats)
library(glue)
library(readxl)
# library(DT)
`%nin%` <- negate(`%in%`)

```


```{r import, echo = F, include=F}
path_ad_loci <- "results/adgwas_loci.csv"
path_adgwas_meta <- "intermediate/gwas_metadata.csv"
path_supp_wightman <- "resources/Wightman2020/41588_2021_921_MOESM4_ESM.xlsx"
path_supp_bellenguez <-
  "resources/Bellenguez2022/41588_2022_1024_MOESM4_ESM.xlsx"

#Export
path_table_s1 <- "docs/TableS1.csv"

if (exists("snakemake")) {
  path_supp_bellenguez <- snakemake@input[["supp_bellenguez"]]
  path_supp_wightman <- snakemake@input[["supp_wightman"]]
  path_ad_loci <- snakemake@input[["ad_loci"]]
  path_adgwas_meta <- snakemake@input[["ad_gwas_meta"]]
  path_table_s1 <- snakemake@params[["TableS1"]]
}

raw_ad_loci <- read_csv(path_ad_loci)
adgwas_meta <- read_csv(path_adgwas_meta)
```


```{r, echo = F}
ad_loci <- raw_ad_loci %>%
  filter(study %nin% c("Jonsson", "Reiman")) %>%
  select(locus, cytoband, locus_ld, Gene = gencode_gene, SNP, CHR, BP,
         A1, A2, FRQ = global_af, OR, P, study) %>%
  mutate(
    FRQ = signif(FRQ, 3),
    OR = signif(OR, 3),
  )

write_csv(ad_loci, path_table_s1)
```

```{r}

bst20_raw <- readxl::read_xlsx(path_supp_bellenguez, sheet = 21, skip = 2) %>%
  janitor::clean_names()

bst20 <- bst20_raw %>%
  select(locus, gene, gene_prioritization_tier) %>%
  filter(gene_prioritization_tier %in% c("Tier 1", "Tier 2"))

wst11 <- readxl::read_xlsx(path_supp_wightman, sheet = 11, skip = 1) %>%
  janitor::clean_names()

bst20 %>%
   filter(gene_prioritization_tier == "Tier 1") %>%
   print(n = Inf)

```


```{r, echo=F, include=F}
## Distinct loci across all studies
n_distinct(ad_loci$locus)
n_distinct(ad_loci$locus_ld)

## Old loci priort to Bellenguez and Wightman
known_loci <- ad_loci %>%
  filter(study %nin% c("Bellenguez", "Wightman"))

n_distinct(known_loci$locus)

## 42 novel loci discovered by Bellenguez and Whightman
novel_loci <- ad_loci %>%
  filter(study %in% c("Bellenguez", "Wightman")) %>%
  anti_join(known_loci, by = "locus") %>%
  # arrange(P) %>%
  # group_by(study) %>%
  distinct(locus, .keep_all = TRUE)

nrow(novel_loci)
novel_loci %>% count(study)

# Number of loci in B & W
n_loci <- ad_loci %>%
  filter(study %in% c("Bellenguez", "Wightman")) %>%
  group_by(locus)

n_distinct(n_loci$locus)

# Number of loci per study
n_loci_study <- ad_loci %>%
  group_by(study)  %>%
  distinct(locus) %>%
  ungroup() %>%
  count(study)

n_loci_study

# Number of independent variants
n_variants <- ad_loci %>%
  filter(study %in% c("Bellenguez", "Wightman")) %>%
  group_by(locus_ld)

n_distinct(n_variants$locus_ld)

# Number of independent variants per study
n_variants_study <- ad_loci %>%
  group_by(study)  %>%
  distinct(locus_ld) %>%
  ungroup() %>%
  count(study)

n_variants_study

```

## Genetic loci associated with Alzheimer’s disease

Wightman et al expanded the sample size of their GWAS to `r filter(adgwas_meta, study == 'Wightman') %>% pull(n)` 
people (`r `filter(adgwas_meta, study == 'Wightman') %>% pull(n_ca)` cases, 
`r NULL` proxy-cases, `r filter(adgwas_meta, study == 'Wightman') %>% pull(n_cn)` controls/proxy-controls). 


Bellenguez et al expanded their earlier GWAS to `r filter(adgwas_meta, study == 'Bellenguez') %>% pull(n)` 
people (`r filter(adgwas_meta, study == 'Bellenguez') %>% pull(n_ca)` cases, 
`r filter(adgwas_meta, study == 'Bellenguez') %>% pull(n_pca)` proxy cases, and 
`r filter(adgwas_meta, study == 'Bellenguez') %>% pull(n_cn)` controls/proxy-controls)


Together, Wightman et al and Bellenguez et al identified `r n_distinct(n_variants$locus_ld)` independent variants 
across `r n_distinct(n_loci$locus)` loci 
(`r filter(n_loci_study, study == "Wightman") %>% pull(n)` and `r filter(n_loci_study, study == "Bellenguez") %>% pull(n)` 
respectively) associated with AD, 34 of which were identified by both studies and `r nrow(novel_loci)` which are novel loci.


With GWAS collectively discovering `r n_distinct(ad_loci$locus)` loci 
(`r n_distinct(ad_loci$locus_ld)` independent variants) associated with AD/dementia 
in European ancestry populations (Figure 2), how many more novel loci remain to be discovered?


```{r, echo = F}
DT::datatable(ad_loci, rownames = FALSE, filter = "top")
```








































